---
title: "Lab 7 Spatial data"
author: "Narjes Mathlouthi"
date: "2/17/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)

#simple feature package/vector data
library(sf)
#tmap interactive map
library(tmap)
```

## Part 1: Spatial data wrangling, visualization

##A. California county outlines(polygons)

```{r}
ca_counties_sf <- read_sf(here("data/ca_counties/CA_Counties_TIGER2016.shp"))

#clean subset

ca_subset_sf <- ca_counties_sf %>% 
  
  janitor::clean_names() %>% 
  select(county_name = name, land_area = aland)

# make a data frame

ca_counties_df <- ca_counties_sf %>% 
  as.data.frame() %>% 
  select(-geometry)
  
```

### Examine and set CRS

CRS : coordinate reference system
we nee

```{r}
ca_subset_sf %>%  st_crs()

ca_subset_sf %>%  raster :: crs()

```


```{r}
ggplot(data = ca_subset_sf)+
  #geom recognizes where to get the spatial info, the geometruy column defines all the characteritics of the map below
  geom_sf(aes(fill = land_area), color = "white", size = 0.1)+
  theme_void() +
  scale_fill_gradientn(colors = c('cyan', 'blue', 'purple'))
```
```{r}
sesbania_sf <- read_sf(here("data/red_sesbania", "ds80.shp")) %>% 
  janitor::clean_names()
  
  
  sesbania_sf %>%  st_crs()
  sesbania_sf %>%  raster::crs()
  
  ### if you know the ESPG
  sesbania_3857_sf<- st_transform(sesbania_sf,3857)
  
  #if you dont know ESPG 
  sesbania_3857_2 <- st_transform(sesbania_sf, st_crs(ca_subset_sf))
  
  #check new one
  sesbania_3857_sf %>%  st_crs()
```

### Plot them together 

```{r}
ca_sesb_sf <-  ca_subset_sf %>% 
  st_join(sesbania_3857_sf)

head(ca_sesb_sf)


sesb_counts_counts_sf <- ca_sesb_sf %>% 
  group_by(county_name) %>% 
  summarize(n_records = sum (!is.na(id)))

head(sesb_counts_counts_sf)
  
```
```{r}
ggplot(data = sesb_counts_counts_sf) + 
  geom_sf(aes(fill = n_records), color = "white", size = 0.1) +
  scale_color_gradientn(colors= c('lightgrey','orange','red')) +
  theme_minimal() +
  labs(fill = "Number of S. punicea records")

```

```{r}

solano_sesb_sf <- sesbania_3857_sf %>% 
  filter(county == "Solano")

solano_sf <- ca_subset_sf %>% 
  filter(county_name == "Solano")

ggplot()+
  geom_sf(data = solano_sf) +
  geom_sf(data = solano_sesb_sf, color = "red")
  
```



```{r}
# Make a map and fill the polygon 
tmap_mode(mode = "view")
tm_shape (ca_subset_sf) +
  tm_fill("land_area",palette ="BuGn") +
  tm_borders(col = "black") +
  tm_shape(sesbania_3857_sf) +
  
  tm_dots()
```

### check out all the palettes:
# tmaptools::palette_explorer() ### need package shinyjs


Let's just look at depth to groundwater (ft) in and around San Joaquin County. 

```{r}
# Just get sj_county
sj_county_sf <- ca_counties_sf %>% 
  filter(name == "San Joaquin")
# Now keep observations for groundwater depth within that county:
sj_depth_sf <- ca_dgw_sf %>% 
  st_intersection(sj_county_sf)
# Make a quick plot of those: 
ggplot() +
  geom_sf(data = sj_county_sf) +
  geom_sf(data = sj_depth_sf, aes(color = dgbs)) +
  theme_void() +
  scale_color_viridis_c()
```

### Check for duplicate locations:

Interpolation chokes when observations at identical location - spatial singularity!

```{r}
well_duplicates <- sj_depth_sf %>% 
  janitor::get_dupes(latitude, longitude)
well_duplicates
# This will cause problems (0 spatial variance). We're just going to get rid of these problematic wells recorded at the same location: 
sj_depth_sf <- sj_depth_sf %>% 
  filter(!local_well %in% well_duplicates$local_well)
# Check again for duplicates: 
sj_depth_sf %>% 
  janitor::get_dupes(latitude, longitude) 
# No more duplicates
```

## Make a variogram: 

```{r}
# Create the variogram
sj_dgw_vgm <- gstat::variogram(dgbs ~ 1, data = sj_depth_sf)
# Plot it
plot(sj_dgw_vgm) # Looks variogramish! Increasing variance as observations get further apart. Let's model this...
# Now create the variogram model, & see ?vgm for other model options: 
sj_dgw_vgm_fit <- fit.variogram(sj_dgw_vgm, 
                                model = vgm(model = "Gau")) ### default starting
sj_dgw_vgm_fit <- fit.variogram(sj_dgw_vgm, 
                                model = vgm(model = "Gau", ### (leave with defaults first)
                                            nugget = 20, sill = 3000, range = 30000))
# Plot them together:
plot(sj_dgw_vgm, sj_dgw_vgm_fit) 
# You can check the sum of squares of residuals for each: 
attr(sj_dgw_vgm_fit, 'SSErr') # ...and you could compare to other models...
```

## Spatial kriging

```{r}
# Make a grid over which we'll krige: 
sj_grid <- st_bbox(sj_county_sf) %>%
  st_as_stars(dx = 1000, dy = 1000) %>%
  st_set_crs(st_crs(sj_county_sf)) %>%
  st_crop(sj_county_sf)
sj_grid
# Look at the grid we made: 
plot(sj_grid)
# Perform ordinary kriging: 
sj_dgw_krige <- krige(dgbs ~ 1, sj_depth_sf, sj_grid, model = sj_dgw_vgm_fit)
# Initial plot of kriging outcome: 
plot(sj_dgw_krige)
# Convert it to a spatial data frame
krige_df <- as.data.frame(sj_dgw_krige) %>% 
  st_as_sf(coords = c("x","y")) %>% 
  drop_na(var1.pred)
st_crs(krige_df) <- 3857
# Then we can use ggplot: 
ggplot(data = krige_df) +
  geom_sf(aes(color = var1.pred)) +
  scale_color_gradient(low = "blue", high = "yellow")
  
```

...& there are ways to convert this to raster data / aggregate if you want...

## End variogram & kriging demo
